Tulai

### 2023-09-22 paper/paper_tulai.qmd
Major edits to remove the google sheet and replace it with a local CSV. Recomputed the values that were in the google sheet using the raw lithic data. Removed many tables that are not mentioned in the manuscript.

---
title: ""
author: ""
date: now
format: 
  html:
    self-contained: true
execute: 
  warning: false
  fig-dpi: 600
title-block-published: "last updated"
editor: visual
---

@@

Cleaning Data:

@@

@@

             TRUE ~ as.character(depth)
  )) %>%

  separate(depth, 
           into = c("upper",
                    "lower"),
           sep = "-",
           remove = FALSE,
           convert = TRUE) %>% 
  mutate(lower = parse_number(lower)) %>% 
  rowwise() %>% 
  mutate(midpoint = mean(c(upper, lower))) %>%
# create 'chronological unit' column
  mutate(chronological_unit = case_when(
      area == "tp1" & depth == "0-10 cm" ~ "c",
      area == "tp1" & depth == "0-20 cm" ~ "c",
      area == "tp1" & depth == "10-30 cm" ~ "c",
      area == "tp1" & depth == "10-90 cm" ~ "c",
      area == "tp1" & depth == "20-30 cm" ~ "c",
      area == "tp1" & depth == "30-40 cm" ~ "c",
      area == "tp1" & depth == "40-50 cm" ~ "c",
      area == "tp1" & depth == "50-100 cm" ~ "c",
      area == "tp1" & depth == "50-60 cm" ~ "b",
      area == "tp1" & depth == "50-70 cm" ~ "b",
      area == "tp1" & depth == "80-90 cm" ~ "b",
      area == "tp1" & depth == "90-100 cm" ~ "b",
      area == "tp1" & depth == "100-110 cm" ~ "b",
      area == "tp1" & depth == "110-120 cm" ~ "b",
      area == "tp1" & depth == "120-130 cm" ~ "b",
      area == "tp1" & depth == "120-140 cm" ~ "b",
      area == "d1" & depth == "0-20 cm" ~ "c",
      area == "d1" & depth == "20-30 cm" ~ "c",
      area == "d1" & depth == "40-60 cm" ~ "a",
      area == "d1" & depth == "50-60 cm" ~ "a",
      area == "d1" & depth == "60-70 cm" ~ "a",
      area == "d1" & depth == "70-80 cm" ~ "a",
      area == "ta" & depth == "0-30 cm" ~ "e",
      area == "ta" & depth == "30-40 cm" ~ "e",
      area == "ta" & depth == "40-50 cm" ~ "e",
      area == "ta" & depth == "50-60 cm" ~ "e",
      area == "ta" & depth == "60-70 cm" ~ "e",
      area == "ta" & depth == "70-120 cm" ~ "d",
      area == "ta" & depth == "120-150 cm" ~ "d",
      area == "ta" & depth == "150-180 cm" ~ "d",
      area == "ta" & depth == "180-190 cm" ~ "d",
      area == "ta" & depth == "190-200 cm" ~ "d",
      area == "ta" & depth == "200-210 cm" ~ "d",
      area == "ta" & depth == "210-220 cm" ~ "d",
      area == "ta" & depth == "220-230 cm" ~ "d",
      area == "ta" & depth == "230-240 cm" ~ "d",
      area == "ta" & depth == "230-250 cm" ~ "d",
      area == "ta" & depth == "250-260 cm" ~ "d",
      area == "ta" & depth == "260-270 cm" ~ "d",

@@

### Summary tables: AREA

@@

# SS: we mainly consider ta, tp1, and d1.
knitr::kable(summary_table_area,
             caption = "Summary of lithics in all areas")
```

@@

knitr::kable(summary_table_area_filtered,
             caption = "Summary of lithics in the main areas")

@@

### Summary tables: chronological unit

```{r}
# Create a summary table
summary_table_unit <- tl_final %>%
  group_by(chronological_unit) %>% 
  summarise(
    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

```{r}
# Generate the table
kable(summary_table_unit, 
      caption = "Summary of lithics by Chronological Units", 
      align = 'c')

### Summary tables: AREA-UNIT:

```{r}
# Create a summary table based on the 'area_unit' column
summary_table_au <- tl_final %>%
  group_by(area_unit) %>%  
  summarise(
    `Cores (n)` = sum(!is.na(core)),  
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

```{r}
# Generate the table

kable(summary_table_al, 
      caption = "Summary of lithics by Area and chronological unit")

### Summary tables: Techno-Typology

# Prepare data for Summary table-area
summary_table_area <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  group_by(area) %>%
  summarise(
    blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
    total = n()
  )

summary_table_area <- summary_table_area %>%
  add_row(
    area = "Total",
    blade_count =              sum(.$blade_count, na.rm = TRUE),
    flake_count =              sum(.$flake_count, na.rm = TRUE),
    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
    tool_count =               sum(.$tool_count, na.rm = TRUE),
    core_count =               sum(.$core_count, na.rm = TRUE),
    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),

# Adding a new row with total counts
summary_table_area <- summary_table_area %>%
  add_row(
    area = "Total",
    blade_count =              sum(.$blade_count, na.rm = TRUE),
    flake_count =              sum(.$flake_count, na.rm = TRUE),
    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
    tool_count =               sum(.$tool_count, na.rm = TRUE),
    core_count =               sum(.$core_count, na.rm = TRUE),
    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
  )

```{r}
# Prepare data for Summary table-area_unit(au)
summary_table_au <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>% 
  group_by(area_unit) %>% 
  summarise(
    blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
    total = n()
  )

# Adding a new row with total counts
summary_table_al <- summary_table_al %>% 
  add_row(
    area_unit = "Total",
    blade_count =              sum(.$blade_count, na.rm = TRUE),
    flake_count =              sum(.$flake_count, na.rm = TRUE),
    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
    tool_count =               sum(.$tool_count, na.rm = TRUE),
    core_count =               sum(.$core_count, na.rm = TRUE),
    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
    total =                    sum(.$total, na.rm = TRUE)  
  )

# rev(sort(table(tl_final$tooltypegroup)))
```

```{r}
#explore tool type group

summary_table_tooltype <- tl_final %>%
  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
  group_by(tooltypegroup) %>% 
  summarise(
  )
```

```{r}
# generate a table

summary_table_al <- summary_table_al %>%
  filter(!(area_unit %in% c("D1_NA", "TP1_NA")))

kable(
  summary_table_al,
  col.names = c(
    "Area-unit",
    "Blade",
    "Flake",
    "Bladelet",
    "Tools",
    "Cores",
    "Core Rejuvenations",
    "Retouched Tools",
    "Utilized Tools",
    "Sickle Shine",
    "Total"
  ),
  caption = "Summary of lithics by techno-typological classification")

@@

```{r}
# Create the summary table based on chronological_unit
summary_table1_unit <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  group_by(chronological_unit) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    numb_cores = sum(!is.na(`core-typology`), na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics) * 100, 2)
  )

@@

```{r}
# Create a summary table based on the 'depth', 'area', and 'chronological unit' columns
tl_final <- tl_final %>%
  mutate(depth = gsub("cm", "", depth))

summary_table_depth_area_unit <- tl_final %>%
  group_by(TA, depth, area, chronological_unit, grouped_level, upper, lower) %>%  
  summarise(
    `Cores (n)` = sum(!is.na(core)),  
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

summary_table_depth_area_unit_to_join <- 
  summary_table_depth_area_unit

# Generate the table
kable(summary_table_depth_area_unit, 
      caption = "Summary of lithics by Depth, Area, and Chronological Unit")
```

```{r}
# Generate a summary table by grouping by 'depth' and 'area'
summary_table_depth_tool_type <- tl_final %>%
  group_by(area, depth) %>%
  summarise(tool_type = n_distinct(tooltypegroup, na.rm = TRUE), .groups = 'drop')

# Create a full summary table by joining it back to the original summary table based on 'depth' and 'area'
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  left_join(summary_table_depth_tool_type, by = c("area", "depth"))

# Remove the existing 'tool_type' if it exists
if ("tool_type" %in% names(summary_table_depth_area_unit)) {
  summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
    select(-tool_type)
}

# Join again based on 'depth', this time being specific about what we want
summary_table_depth_area_unit <- 
  summary_table_depth_area_unit %>%
  left_join(summary_table_depth_tool_type %>% select(depth, tool_type), 
            by = "depth")

# Generate the table

kable(summary_table_depth_area_unit, 
      caption = "Summary of lithics by Depth, Area, and chronological unit")

```{r}
# Specify the exact columns want to keep
desired_columns <-
  c(
    "TA",
    "depth",
    "area",
    "chronological_unit",
    "grouped_level",
    "upper",
    "lower",
    "Cores (n)",
    "Core rejuvenations pieces (n)",
    "Blade (n)",
    "Flake (n)",
    "Bladelet (n)",
    "Retouched Tools (n)",
    "Utilized (n)",
    "Sickle Shine (n)",
    "tool_type"
  )

# Filter the tl_final dataframe
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  filter(area %in% c('ta', 'd1', 'tp1'))

# Generate the table
kable(summary_table_depth_area_unit, 
      caption = "Summary of lithics by Depth, Area, and chronological_unit")
```

# import table with depths that we have subdivided
depths_divided_tbl <- 
  read_csv(here("data/Lookup Table - Sheet1.csv")) %>% 
  mutate(division_id = rep(c("a", "b"),
                           nrow(.)/2))

# google sheet has 42 rows, and cols for 
# area, chronological_unit, level, depth. What is the equivalent above? tp1_c_3_a

# This 'depths_divided_tbl' needs more information, for example, some depth ranges

# join with lithic data that 
summary_table_depth_area_unit_to_join %>% 
  left_join(depths_divided_tbl) %>% 
  mutate(label = paste(area, 
                       chronological_unit, 
                       # level, # need this
                       division_id,
                       sep = "_")) %>% View

```{r}
#| eval: false

# Here, I've imported the table 'Summary by Depth, Area, chronological_unit' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.

# BM: looks like you need this data starting from around here in the document

library(googlesheets4)

google_sheet_url <- 
  "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"
sum_depth_sheet <- read_sheet(google_sheet_url)

# Create the 'label' column
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))

# Add a suffix only to duplicates
sum_depth_sheet <- sum_depth_sheet %>%
  group_by(label) %>%
  mutate(suffix = case_when(
    n() == 1 ~ NA_character_,
    TRUE ~ letters[row_number()]
  )) %>%
  ungroup()

# Append the suffix to 'label'
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
  select(-suffix)

# Relocate 'label' column after 'level'
sum_depth_sheet <- sum_depth_sheet %>%
  relocate(label, .after = level)

# BM: from my perspective, what would be ideal here is to have the 
# excavation area and volume calculations done here in R. I feel anxious about
# the possibilities of typos when these values are calculated somewhere I can't 
# see and pasted into the google sheet. I prefer for the workflow of calculations
# to be as transparent and traceable as possible. From this point in the qmd I
# think we can use the volume values from this google sheet. Another option is to 
# export the google sheet as a CSV, and put that in the data folder in this project. 
# Then we can use read_csv to import it.

#SS: is that true?: (I have a weird number in row 5, cl Depth and couldn't find out what happened)

#### previously I had a column'excavation area'; I seprated them in 2 columns in order to calculate the colume; but now when I wanted to run the code it said "! object 'excavation area' not found"; would it be OK to delete these lines?

# Separate `excavation area` into `length` and `width` and calculate `volume`
sum_depth_sheet <- sum_depth_sheet %>%
  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
  mutate(
    length = as.numeric(length),
    width = as.numeric(width),
    `thickness` = `thickness` / 100  # Convert from cm to m
  ) %>%
  mutate(
    volume = length * width * `thickness`  # Calculate volume in m^3
  )

# Calculate lithic volumetric density 
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    `lithic_dens` = round(total_lithics / volume, 2)
  )

# Calculate retouch frequency 
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    retouch_freq = round(retouch_tool / volume, 2)  
  )

#try with summary_table_depth_area_unit
```{r}
# Create the 'label' column
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  mutate(label = paste(area, chronological_unit, grouped_level, sep = "_"))

# Add a suffix only to duplicates
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  group_by(label) %>%
  mutate(suffix = case_when(
    n() == 1 ~ NA_character_,
    TRUE ~ letters[row_number()]
  )) %>%
  ungroup()

# Append the suffix to 'label'
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
  select(-suffix)

# Relocate 'label' column after 'level'
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  relocate(label, .after = grouped_level)

------------------------------------------
# working up to here. I need to add length, width, and thickness columns
------------------------------------------

# Separate `excavation area` into `length` and `width` and calculate `volume`
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
  mutate(
    length = as.numeric(length),
    width = as.numeric(width),
    `thickness` = `thickness` / 100  # Convert from cm to m
  ) %>%
  mutate(
    volume = length * width * `thickness`  # Calculate volume in m^3
  )

# Calculate lithic volumetric density 
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  mutate(
    `lithic_dens` = round(total_lithics / volume, 2)
  )

# Calculate retouch frequency 
summary_table_depth_area_unit <- summary_table_depth_area_unit %>%
  mutate(
    retouch_freq = round(retouch_tool / volume, 2)  
  )
```

```{r}
library(ggpmisc)

# Create the plot: WABI_Level_logscale
# here I do not consider those levels with a value 0 of retouch frequency. 
filtered_data <- sum_depth_sheet %>% 
  filter(lithic_dens > 0 & retouch_freq > 0)

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  geom_text(aes(label = label), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 4) +
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

# Print the plot to display it
print(WABI_Level)

```{r}
#PCA_Level: preparing data

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should I consider the thickness?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
  ##core frequency
  ##blank frequency
  ##retouch frequency
  ##tool diversity
  ##thickness of deposit (???)
  ##lithic density

#Calculate Core Frequency
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    core_freq = round(cores / volume, 2)  
  )

# calculate Blank Frequency
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    blank_freq = round((flake + blade + bladelet) / volume, 2)  
  )

# Calculate tool diversity using Menhinick's index
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    total_tools = retouch_tool + utilized + sickle_shine,
    tool_diversity = ifelse(total_tools > 0, n_distinct(tool_type) / sqrt(total_tools), NA)
  )

```{r}
# performin PCA

@@

sum_depth_sheet_subset <- 
 sum_depth_sheet %>%
  drop_na(tool_diversity) #%>% 
  #filter(!label %in% c("tp1_c_3_a",
 #                      "d1_a_2_b",
  #                     "d1_c_3_b"))

pca_data <- 
sum_depth_sheet_subset %>% 
  select(lithic_dens, 
         core_freq, 
         blank_freq, 
         retouch_freq, 
         tool_diversity)

@@

# View the summary results
summary(res.pca)

```{r}
#Plot: Contribution of variables for each of the four relevant PCA dimensions

#save contribution plots into one objects

@@

#combine individual plots into one plot 
library(cowplot)
plot_grid(plotlist = contribution_plots,
                   ncol = 2)

```{r}
# Extract contributions of variables to each principal component
contributions <- res.pca$var$contrib

# Compute average contribution for each dimension
avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)
avg_contrib_dim5 <- mean(contributions[,5], na.rm = TRUE)

# Find variables that contribute more than average to each dimension
high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]
high_contrib_dim5 <- contributions[contributions[,5] > avg_contrib_dim4, 5]

# Sort them in descending order
sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)
sorted_high_contrib_dim5 <- sort(high_contrib_dim5, decreasing = TRUE)

# Print or use for further analysis
print("Significant contributions to Dim1:")
print(sorted_high_contrib_dim1)

print("Significant contributions to Dim2:")
print(sorted_high_contrib_dim2)

print("Significant contributions to Dim3:")
print(sorted_high_contrib_dim3)

print("Significant contributions to Dim4:")
print(sorted_high_contrib_dim4)

print("Significant contributions to Dim5:")
print(sorted_high_contrib_dim5)
```

@@

```{r}
# Inspect eigenvalues
res.pca$eig
```

>>>>

---
title: ""
author: ""
date: now
format: 
  html:
    self-contained: true
execute: 
  warning: false
  fig-dpi: 600
title-block-published: "last updated"
---

@@

Import the data

@@

# this file is prepared from the google sheet here
# https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0
# and simplified using code in the scraps.R file
excavation_data <- read_csv(here("data/excavation_data.csv"))
```

Clean the data

@@

             TRUE ~ as.character(depth)
  )) %>% 
  separate(depth, 
           into = c("upper",
                    "lower"),
           sep = "-",
           remove = FALSE,
           convert = TRUE) %>% 
  mutate(lower = parse_number(lower)) %>% 
  rowwise() %>% 
  mutate(midpoint = mean(c(upper, lower))) %>%
# create 'chronological unit' column
  mutate(chronological_unit = case_when(
      area == "tp1" & depth == "0-10 cm" ~ "c",
      area == "tp1" & depth == "0-20 cm" ~ "c",
      area == "tp1" & depth == "10-30 cm" ~ "c",
      area == "tp1" & depth == "10-90 cm" ~ "c",
      area == "tp1" & depth == "20-30 cm" ~ "c",
      area == "tp1" & depth == "30-40 cm" ~ "c",
      area == "tp1" & depth == "40-50 cm" ~ "c",
      area == "tp1" & depth == "50-100 cm" ~ "c",
      area == "tp1" & depth == "50-60 cm" ~ "b",
      area == "tp1" & depth == "50-70 cm" ~ "b",
      area == "tp1" & depth == "80-90 cm" ~ "b",
      area == "tp1" & depth == "90-100 cm" ~ "b",
      area == "tp1" & depth == "100-110 cm" ~ "b",
      area == "tp1" & depth == "110-120 cm" ~ "b",
      area == "tp1" & depth == "120-130 cm" ~ "b",
      area == "tp1" & depth == "120-140 cm" ~ "b",
      area == "d1" & depth == "0-20 cm" ~ "c",
      area == "d1" & depth == "20-30 cm" ~ "c",
      area == "d1" & depth == "40-60 cm" ~ "a",
      area == "d1" & depth == "50-60 cm" ~ "a",
      area == "d1" & depth == "60-70 cm" ~ "a",
      area == "d1" & depth == "70-80 cm" ~ "a",
      area == "ta" & depth == "0-30 cm" ~ "e",
      area == "ta" & depth == "30-40 cm" ~ "e",
      area == "ta" & depth == "40-50 cm" ~ "e",
      area == "ta" & depth == "50-60 cm" ~ "e",
      area == "ta" & depth == "60-70 cm" ~ "e",
      area == "ta" & depth == "70-120 cm" ~ "d",
      area == "ta" & depth == "120-150 cm" ~ "d",
      area == "ta" & depth == "150-180 cm" ~ "d",
      area == "ta" & depth == "180-190 cm" ~ "d",
      area == "ta" & depth == "190-200 cm" ~ "d",
      area == "ta" & depth == "200-210 cm" ~ "d",
      area == "ta" & depth == "210-220 cm" ~ "d",
      area == "ta" & depth == "220-230 cm" ~ "d",
      area == "ta" & depth == "230-240 cm" ~ "d",
      area == "ta" & depth == "230-250 cm" ~ "d",
      area == "ta" & depth == "250-260 cm" ~ "d",
      area == "ta" & depth == "260-270 cm" ~ "d",

@@

### Summary tables of the lithic assemblages

#### Basic typological classes by excavation area

@@

@@

# BM: this is table 1 in the manuscript
knitr::kable(summary_table_area_filtered,
             caption = "Summary of lithics in the main areas")

@@

#### Techno-typological classes by excavation area

@@

### Combine lithic data with excavation data to prepare for WABI

```{r}
# Create the 'label' column
tl_final_depth_area_unit <- 
  tl_final %>%
  mutate(label1 = paste(area, 
                       chronological_unit, 
                       grouped_level, 
                       sep = "_")) %>% 
  mutate(label2 = paste(area, 
                       chronological_unit, 
                       grouped_level, 
                       upper,
                       lower,
                       sep = "_")) %>% 
  # we don't care about lithics with missing and incomplete 
  # excavation provenance, so drop those
 filter(!str_detect(label1, "NA")) %>% 
  # so we can join this column ok 
  mutate(grouped_level = parse_number(grouped_level))

# BM: For the individual lithic data, what area/chrono-unit/grouped level 
# combinations do we have? We only want to match those with the 
# excavation data that we can calclate volume with.

lithics_unique_labels <- 
tl_final_depth_area_unit %>% 
  pull(label2) %>% 
  unique() %>% 
  sort()

# BM: prepare excavation data here
excavation_data_volume <-
  excavation_data %>% 
  mutate(thickness_of_excavation_unit  = lower_depth - upper_depth) %>% 
  mutate(excavation_volume = excavation_dimension_1 * 
           excavation_dimension_2 * 
           thickness_of_excavation_unit / 100)

# BM: for the WABI plot, we have our primary analytical unit as
# excavation_area / chronological_unit / depth which is our label2
# variable, each row in the excavation_data_volume table is a 
# unique combination of excavation_area / chronological_unit / depth

# BM: Now we subset the excavation data so we only look at the 
# data that we need for the lithics, ie. we drop excavation data 
# that has no corresponding lithics

tl_final_excavation_data_volume  <- 
excavation_data_volume %>% 
  filter(label2 %in% lithics_unique_labels) %>% 
  # now we join the excavation data to the individual lithic records
  right_join(tl_final_depth_area_unit)

# get the counts of artefact types per ex-area/chrono-unit/depth

tl_final_excavation_data_volume_artefact_counts <- 
tl_final_excavation_data_volume %>% 
  group_by(label2) %>% 
   summarise(
    blade_count =              sum(blank == "blade", na.rm = TRUE),
    flake_count =              sum(blank == "flake", na.rm = TRUE),
    bladelet_count =           sum(blank == "bladelet", na.rm = TRUE),
    tool_count =               sum(retouch == 1, 
                                   na.rm = TRUE) + sum(utilization == 1, 
                                                       na.rm = TRUE),
    core_count =               sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count =  sum(!is.na(`corefrag`), 
                                   na.rm = TRUE),
    retouched_tool_count =     sum(retouch == 1, 
                                   na.rm = TRUE),
    utilized_tool_count =      sum(utilization == 1, 
                                   na.rm = TRUE),
    shine_tool_count =         sum(`sickle shine` == 1, 
                                   na.rm = TRUE),
    tool_type_count = n_distinct(tooltypegroup),
    excavation_volume = unique(excavation_volume)

  ) %>% 
 # BM:  "total lithics" is everything except sickle shine, 
 # "retouch frequency" is retouched_tool_count / "total lithics"
  rowwise() %>% 
  mutate(total_lithics = sum(blade_count,
                             flake_count,
                             bladelet_count,
                             tool_count,
                             core_count,
                             core_rejuvenation_count,
                             retouched_tool_count,
                             utilized_tool_count)) %>% 
  mutate(retouch_frequency =  retouched_tool_count / total_lithics)

# BM: get the total number of artefacts, number of retouched pieces, and 
# artefact density. I think this gets us to something similar to the google
# sheet, the values are not exact, but I have no way to check how how
# calculated the values in the google sheet so 
sum_depth_sheet <- 
tl_final_excavation_data_volume_artefact_counts %>% 
  # some ex-area/chrono-unit/depth have no volume, so drop those
  drop_na(excavation_volume) %>% 
  mutate(lithic_density = total_lithics / excavation_volume)
```

### Plot the WABI

```{r}
#| label: fig-wabi
#| fig-cap: "Relationship between Retouch Frequency and Lithic Volumetric Density"

library(ggpmisc)
library(ggrepel)

# Create the plot: WABI_Level_logscale
# here I do not consider those levels with a value 0 of retouch frequency. 
filtered_data <- 
  sum_depth_sheet %>% 
  filter(lithic_density > 0 & retouch_frequency > 0)

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_density,
                y = retouch_frequency)) +
  geom_point(color = "black", 
             size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  geom_text_repel(aes(label = label2)) +
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 4) +
  labs(
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

# Print the plot to display it
WABI_Level

@@

### Compute PCA

```{r}
# PCA_Level: preparing data

# PCA_level: for PCA Bicho and Cascalheira considered these variables: 
# - estimated area (as we calculated based on levels, should I consider the thickness?),
# - core frequency: core_count / total_lithics
# - blank frequency: (flake + blade + bladelet) / total_lithics
# - chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), 
# - feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), 
# - retouch frequency: retouch_frequency
# - tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index) BM: tool diversity from B & C: "calculated using the Menhinick’s index in which the number of tool types represented is divided by the square root of the total number of retouched tool".

sum_depth_sheet_for_pca <- 
 sum_depth_sheet %>%
  mutate(
      core_frequency = core_count / total_lithics,
      blank_frequency = (flake_count + blade_count + bladelet_count) / total_lithics,
      total_tools = retouched_tool_count + utilized_tool_count + shine_tool_count,
    # Calculate tool diversity using Menhinick's index
      tool_diversity = tool_type_count / sqrt(total_tools)) %>% 
   # drop rows that have no tools 
  filter(total_tools != 0)

# performing PCA

@@

sum_depth_sheet_subset <- 
 sum_depth_sheet_for_pca %>%
  drop_na(tool_diversity) #%>% 
  # filter(!label %in% c("tp1_c_3_a",
  #                      "d1_a_2_b",
  #                     "d1_c_3_b"))

pca_data <- 
sum_depth_sheet_subset %>% 
  select(lithic_density, 
         core_frequency, 
         blank_frequency, 
         retouch_frequency, 
         tool_diversity)

@@

# View the summary results
# summary(res.pca)
```

```{r}
#| label: pca-contrib
#| fig-cap: "PCA contribution plots"
#| fig-height: 10

# Plot: Contribution of variables for each of the four relevant PCA dimensions
# save contribution plots into one objects

@@

#combine individual plots into one plot 
library(cowplot)
plot_grid(plotlist = contribution_plots,
                   ncol = 2)
```

@@

```{r}
# Inspect eigenvalues

eigen_df <- data.frame(
   eigenvalue = res.pca$eig[,1],
  `percentage of variance` = res.pca$eig[,2],
  `cumulative percentage of variance` = res.pca$eig[,3]
)

library(kableExtra)
